---
title: "R for ETL"
author: "Steph Locke"
date:  "`r Sys.Date()`"
output: 
  revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, dev="svg",
     out.height = "800px",  strip.white=TRUE
     ,collapse = TRUE, dev.args = list(bg="transparent"), warning = FALSE
)
library(tidyverse)
```

## Agenda
- Data import
- Data transformation
- Data validation
- Data export
- Framework
- Infrastructure


## Steph Locke & Locke Data
- Steph
    + MVP
    + Learn R books: [geni.us/rfundamentals](//geni.us/rfundamentals)
- Locke Data
    + Consultancy focussed on helping organisations get started with data science

- @stefflocke @lockedata
- steph@itsalocke.com
- itsalocke.com

# Data import
## Key points
- Many, many packages for data access
- Bring Your Own Package (BYOP)
- Build programmatic connections
- Sophisticated error handling

## R packages
- DBI
- odbc
- jsonlite
- xml2
- httr
- feather
- haven
- readr
- readxl
- googlesheets
- tabulizer

## Example - basic CSV read
```{r simpleread}
"data/lego-database/colors.csv" %>% 
  read_csv() ->
  colours
```

## Example - more complex CSV read
```{r complexcsvvars}
file = "data/lego-database/colors.csv" 
colnames = c("ID","NAME","HEX","TRANSPARENT")
datatypes = cols(ID=col_integer(), .default= col_character())
```

```{r complexcsv}
file %>% 
  read_csv(col_names = colnames,
           col_types = datatypes,
           skip = 1) ->
  colours2
```

## Example - CSV read benchmark
```{r benchmark, message=FALSE}
file = "data/world-development-indicators/indicators.csv"
file.size(file)/1024^2 #megabytes

microbenchmark::microbenchmark(
  readr::read_csv(file, progress=FALSE),
  times=10
)
```

## Example - bulk read and combine (setup)
(or a sneakpeak into the write stages!)
```{r  bulkreadprepsetup}
dir = "example/big-read"
dir.create(dir, recursive = TRUE)
```

```{r  bulkreadprep}
library(data.table)
file %>% 
  fread() %>% 
  split(by="CountryCode") %>% 
  map(~fwrite(.,
      file.path(dir,paste0( unique(.$CountryCode),".csv")))) %>% 
  names() %>% 
  head()
```

## Example - bulk read and combine
```{r bulkread, message=FALSE}
dir %>% 
  list.files("*.csv", full.names = TRUE) %>% 
  map_df(read_csv) %>% 
  nrow()
```

## Example - DB connectivity
(sqlite for easy offline work)

```{r dbconn}
db = "data/world-development-indicators/database.sqlite"

library(DBI)
mydb<-dbConnect(RSQLite::SQLite(), db)
dbListTables(mydb)
```

## Example - Push SQL to the DB
(sqlite for easy offline work)
```{r dbplyrdemo, eval=FALSE}
library(dbplyr)
mydb %>% 
  tbl("Country") %>% 
  filter(Region=="South Asia") %>% 
  top_n(10) 
```

## Exercises
1. Read the `lego/inventory_parts.csv` data into memory
2. Connect to the world development indicators database and use `dbGetQuery` to return the country notes and the associated data aboout the countries

# Data transformation
## Key points
- Quicker and smarter than SQL for a lot of stuff
- Lots of packages
- Reproducible
- Control flow

## R packages
- tidyverse (dplyr, purr, stringr, forcats, lubridate)
- tidytext
- fuzzyjoin
- anytime
- glue
- iptools and urltools

## Example - Standardising names
```{r standardisenames}
file %>% 
  read_csv(progress = FALSE) %>% 
  rename_all(tolower) %>% 
  names()
```

## Example - Reshaping data
```{r reshapedata}
file %>% 
  read_csv(progress = FALSE) %>% 
  filter(IndicatorCode=="SP.ADO.TFRT") %>% 
  spread(Year, Value) %>% 
  names()
```

## Example - Fuzzy joins
```{r fuzzyjoin}
library(fuzzyjoin)
colours() %>% 
  data_frame(name=.) ->
  r_colours
colours %>% 
  stringdist_left_join(r_colours, distance_col = "dist")
```

## Example - Cleanse all text data
```{r cleantext}
"data/world-development-indicators/series.csv" %>% 
  read_csv() %>% 
  mutate_if(is.character,
            ~str_replace_all(
              str_trim(
                str_to_lower(.)
                ),
              "[:punct:]",""))
```

## Exercise
1.Load the world development indicators series notes file. Use `str_trunc()` to truncate all text to 126 characters.

# Data validation
## Key points
- Apply validation criteria in pipeline
- Route data based on quality

## R packages
- assertr
- assertive
- pointblank

## Example - Checking business rules
```{r assertrverify, error=TRUE}
library(assertr)
mydb %>% 
  dbReadTable("Indicators") %>% 
  verify(Value > 1)
```

## Example - Relative assertions
```{r assertrinsist, error=TRUE}
library(assertr)
mydb %>% 
  dbReadTable("Indicators") %>% 
  insist(within_n_sds(3), Value)
```

## Exercise
1. Write a check that says the years in the indicators data must be between 1950 and 2015

# Data export
## Key points
- CSV and other interchange formats ftw
- Audit and restartability

## R packages
- DBI
- writexl and openxlsx
- httr
- haven

## Example - Facetted writes
```{r bigwrite}
file %>% 
  fread() %>% 
  split(by="CountryCode") %>% 
  map(~fwrite(.,
      file.path(dir,paste0( unique(.$CountryCode),".csv")))) %>% 
  names() %>% 
  head()
```

## Example - Output everything
```{r bigwriter}
dir = "example/big-writes"
dir.create(dir, recursive = TRUE)
mydb %>% 
  dbListTables() %>% 
  map(~write_csv(dbReadTable(mydb, .),
                 paste0(dir,"/",.,".csv") ))
  
```

## Exercise
1. Write a table from the db to a spreadsheet using the `writexl` package.

# Framework 
## Key points
- Logging
- Reusability
- Secrets
- File locations
- Profiling performance
- Object models
- Testing

## R packages
- here
- profvis
- R6
- reticulate
- V8
- testthat
- tidyeval
- sparklyr

## Example - Testing code
```{r testthatfn}
library(testthat)
general_clean<-function(df){
  df %>% 
    mutate_if(is.character, str_to_lower)
}
```

```{r testthat, error=TRUE}
test_that("clean lowercases",{
  mydf<-data_frame(a=LETTERS[1:5], b=1:5)
  expect_equal(general_clean(mydf),
               data_frame(a=letters[1:5],b=1:5))
  expect_type(general_clean(mydf), "data.frame")
})
```


## Exercise
1. Write a function that scales numeric columns in a table using the `scale()` function
2. Test the function works as expected

# Infrastructure
## Key points
- Out of memory processing
- Email
- Managing change


## R packages
- Microsoft ML
- gmail
- checkpoint
- packrat
- miniCRAN
- git2r
- openssl

## Example - Building your own repository
```{r minicran, eval=FALSE}
library(miniCRAN)
makeRepo("datasauRus")
```


# Conclusion
## Conclusion
- Data import
- Data transformation
- Data validation
- Data export
- Framework
- Infrastructure

## Wrapup
Q&A